How economic, health and social factors affect life expectancy in developing countries?

### 2025-04-10 paper.Rmd
final verision 1

# Contributions [TODO: UPDATE THIS]

@@

```{r}

ggplot(data_interest, aes(x = percentage_expenditure, y = life_expectancy)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  labs(x = "Percentage Expenditure (as percentage of GDP per capita)", y = "Life Expectancy", 
       title = "Percentage Expenditure vs Life Expectancy") +
  theme_minimal()

ggplot(data_interest, aes(x = log1p(percentage_expenditure), y = life_expectancy)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(x = "Log(1 + Percentage Expenditure)", y = "Life Expectancy", 
       title = "Log-Transformed Percentage Expenditure vs Life Expectancy") +
  theme_minimal()

@@

```{r}

@@

```{r}

@@

```{r}

@@

secondary_model_infant_lg = lm(life_expectancy ~ infant_deaths_lg1p + hiv_aids_lg1p + 
                         schooling + percentage_expenditure_lg1p + alcohol, 
                       data=data_interest_final)
summary(secondary_model_infant_lg)

@@

```{r}

# Create the initial data frame with regression results
results <- data.frame(
  Term = c("(Intercept)", "infant_deaths_lg1p", "hiv_aids_lg1p", "schooling", 
           "percentage_expenditure_lg", "alcoholNegligible"),
  Estimate = c(57.86168, -0.37360, -6.22788, 1.06694, 0.54056, -1.51114),
  Std_Error = c(0.72966, 0.06815, 0.13066, 0.05544, 0.06703, 0.23922),
  t_value = c(79.300, -5.482, -47.663, 19.244, 8.064, -6.317),
  p_val = c("< 2e-16", "4.99e-08", "< 2e-16", "< 2e-16", "1.57e-15", "3.57e-10")
)

@@

```{r}
plot(secondary_model_infant_lg, which = 1)
hist(rstandard(secondary_model_infant_lg))
```
### TODO: Add A Couple Sentences about how the model fits the assumptions now

@@

```{r}

>>>>

# Contributions

## Final Contributions

Siddharth: Completed writing for section 3, 4, and 6. Edited section 5. Created an Rmd file, performed regression analysis in R, formatted text in R Studio, generated data visualizations.

Hasan: Completed writing for section 1,2,3 and 6. Conducted research and constructed the research question.

Arshh: Wrote section 5, helped with interpretation of final model results and model transformations, conducted research and literature review, helped formulate research question and helped with formatting.

## Preliminary Contributions

@@

```{r}
#| label: fig-healthexpenditureplots
#| layout-ncol: 2
#| fig-align: "center"
#| warning: false
#| message: false
#| echo: false

ggplot(data_interest, aes(x = percentage_expenditure, y = life_expectancy)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  labs(x = "Health Expenditure", y = "Life Expectancy", 
       title = "Health Expenditure (as percentage of GDP per capita) vs Life Expectancy") +
  theme_minimal()

ggplot(data_interest, aes(x = log1p(percentage_expenditure), y = life_expectancy)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(x = "Log(1 + Health Expenditure)", y = "Life Expectancy", 
       title = "Log-Transformed Health Expenditure vs Life Expectancy") +
  theme_minimal()

@@

```{r}
#| label: fig-infantdeathstransformationplots
#| layout-ncol: 2
#| fig-align: "center"
#| warning: false
#| message: false
#| echo: false

@@

```{r}
#| label: fig-hivtransformationplots
#| layout-ncol: 2
#| fig-align: "center"
#| warning: false
#| message: false
#| echo: false

@@

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false

@@

secondary_model_infant_lg = lm(life_expectancy ~ infant_deaths_lg1p + hiv_aids_lg1p + 
                         schooling + percentage_expenditure_lg1p + alcohol, 
                       data=data_interest_final)
# summary(secondary_model_infant_lg)

@@

```{r}
#| label: tbl-finalcoefficentplots
#| warning: false
#| message: false
#| echo: false

results <- data.frame(
  Term = c("(Intercept)", "log(1 + Infant Deaths)", "log(hiv/aids Deaths + 1)", "schooling", 
           "log(1 + percentage expenditure)", "alcoholNegligible"),
  Estimate = c(57.86168, -0.37360, -6.22788, 1.06694, 0.54056, -1.51114),
  Std_Error = c(0.72966, 0.06815, 0.13066, 0.05544, 0.06703, 0.23922),
  t_value = c(79.300, -5.482, -47.663, 19.244, 8.064, -6.317),
  p_val = c("< 2e-16", "4.99e-08", "< 2e-16", "< 2e-16", "1.57e-15", "3.57e-10")
)

@@

```{r}
#| label: fig-finalmodelassumptionsplot
#| layout-ncol: 2
#| fig-align: "center"
#| warning: false
#| message: false
#| echo: false

plot(secondary_model_infant_lg, which = 1)
hist(rstandard(secondary_model_infant_lg))
```
Our preliminary model violated the assumptions of linearity and constant variance. There was also a minor violation of the normal errors assumption as they were slightly skewed left. However, post adding the log transformations, the improved model does not violate these assumptions.

@@

```{r}
#| label: fig-stepwiseplots
#| layout-ncol: 2
#| fig-align: "center"
#| warning: false
#| message: false
#| echo: false
